---
title: "Final"
output: pdf_document
date: "2024-11-09"
---
# Set Up
```{r Setting up Knitr, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, tidy.opts = list(width.cutoff = 60), tidy = TRUE) 
```

```{r Libraries}
libraries <- c('tidyverse', 'splines', 'rpart', 'rpart.plot', 'randomForest', 'gbm', 'caret', 'janitor', 'survey', 'irr')
for (pkg in libraries){
  library(pkg, character.only = TRUE)
}
```

# Clean Up
The COVID-19 Vaccination Module is available on the Combined Land Line and Cell Phone survey and versions 1-3 for states that conducted more than one version of the questionnaires. 
```{r Importing Files}
brfss <- read_csv("brfss.csv") %>% 
  clean_names()
brfssv1 <- read_csv("v1_2023.csv") %>% 
  clean_names()
brfssv2 <- read_csv("v2_2023.csv") %>% 
  clean_names()
brfssv3 <- read_csv("v3_2023.csv") %>% 
  clean_names()
```

The following states had data on COVID-19 Vaccination status: 

* Combined Land Line and Cell Phone data: Arkansas, Connecticut, Delaware, Georgia, Guam, Hawaii, Illinois, Michigan, Minnesota, Nebraska, New Hampshire, New Jersey, New Mexico, North Carolina, North Dakota, Puerto Rico, Vermont, Virginia, West Virginia, Wisconsin, Wyoming

* Combined Land Line and Cell Phone data, version 1: Oklahoma

* Combined Land Line and Cell Phone data, version 2: Maryland

* Combined Land Line and Cell Phone data, version 3: Maryland
```{r Subsetting and Combining States}
# Subsetting and Combining States
brfss_s <- brfss[brfss$state %in%
c(5, 9, 10, 13, 66, 15, 17, 26, 27, 31, 33, 34, 35, 37, 38, 72, 50, 51, 54, 55, 56), ]
brfss_s <- brfss_s %>% 
  rename(finalwt = llcpwt) %>% 
  rename(finalcpwt = cllcpwt)
brfssv1_s <- brfssv1[brfssv1$state %in% c(40), ]
brfssv1_s <- brfssv1_s %>% 
  rename(finalwt=lcpwtv1) %>% 
  rename(finalcpwt = clcwtv1)
brfssv2_m <- brfssv2[brfssv2$state %in% c(24), ]
brfssv2_m <- brfssv2_m %>% 
  mutate(finalwt=lcpwtv2*0.50) %>% 
  mutate(finalcpwt = clcwtv2*0.50) %>% 
  subset(select=-c(lcpwtv2, clcwtv2))
brfssv3_m <- brfssv3[brfssv3$state %in% c(24), ]
brfssv3_m <- brfssv3_m %>% 
  mutate(finalwt=lcpwtv3*0.50) %>% 
  mutate(finalcpwt = clcwtv3*0.50) %>% 
  subset(select=-c(lcpwtv3, clcwtv3))
brfss_t <- rbind(brfss_s, brfssv1_s, brfssv2_m, brfssv3_m)
```

Only people who had COVID-19 are able to develop long COVID. 
```{r Subsetting Covid Positives}
# Subsetting Covid Positives
brfss_c <- subset(brfss_t, covidpo1==1) 
```


```{r Cleaning and Creating Variables}
# Long Covid - Defining missingness, factoring, creating level names
brfss_c$covidsm1[brfss_c$covidsm1==7|brfss_c$covidsm1==9] <- NA
brfss_c$covidsm1 <- factor(
  ifelse(brfss_c$covidsm1==1, "Pos", "Neg")
)
brfss_c$covidsm1 <- relevel(brfss_c$covidsm1, ref="Pos")
# Defining predictor missingness
## Race
brfss_c$race[brfss_c$race==9] <- NA

## Education
brfss_c$educag[brfss_c$educag==9] <- NA

## Income
brfss_c$incomg1[brfss_c$incomg1==9] <- NA

## Insurance Status
brfss_c$hlthpl1[brfss_c$hlthpl1==9] <- NA

## Smoking
brfss_c$smoker3[brfss_c$smoker3==9] <- NA

# Factoring Predictors
variable <- c("age_g", "sexvar", "race", "educag", "incomg1", "hlthpl1", "smoker3")
for (var in variable) {
brfss_c[[var]] <- factor(brfss_c[[var]])
}

# BMI - Creating a continuous BMI variable, removing outliers
brfss_c$wt <- brfss_c$wtkg3/100
brfss_c$ht <- brfss_c$htm4/100
brfss_c$bmi <- brfss_c$wt/brfss_c$ht^2
iqr <- IQR(brfss_c$bmi, na.rm = TRUE)
qnt <- quantile(brfss_c$bmi, probs=c(.25, .75), na.rm = TRUE)
lower <- qnt[1] - 1.5*iqr
upper <- qnt[2] + 1.5*iqr
brfss_c <- brfss_c %>% 
  subset(brfss_c$bmi>lower & brfss_c$bmi<upper)

# Mental Health - Defining missingness and 0
brfss_c$menthlth[brfss_c$menthlth==77|brfss_c$menthlth==99] <- NA
brfss_c$menthlth[brfss_c$menthlth ==88] <- 0

# COVID-19 Vaccine Status - Creating a factor vaccine variable starting with 0
brfss_c$cvax <-factor(
  ifelse(brfss_c$covidva1==2,0, 
                       ifelse(brfss_c$covidnu2==1,1,
                              ifelse(brfss_c$covidnu2==2,2,
                                     ifelse(brfss_c$covidnu2==3,3,
                                            ifelse(brfss_c$covidnu2==4,4,
                                                   ifelse(brfss_c$covidnu2==5,5, NA))))))
)
```

```{r Addressing Missingness}
# Complete Case Analysis
brfss_c <- na.omit(brfss_c[, c("age_g", "sexvar", "race", "educag", "incomg1", "hlthpl1", "smoker3", "bmi", "menthlth", "cvax", "covidsm1")])
```

There are 6,942 participants with long COVID and 46,886 without. I am creating class weights to address the imbalance. 
```{r Calculating Class Weights}
class_counts <- table(brfss_c$covidsm1)
total_samples <- sum(class_counts)
num_classes <- length(class_counts)
class_weight <- total_samples / (num_classes * class_counts)
class_weight
brfss_c$classwt <- ifelse(brfss_c$covidsm1=="Pos", class_weight["Pos"], class_weight["Neg"])
```

# Machine Learning Models
```{r Logistic Regression}
# Logistic Regression
set.seed(1234)
lc_log <- train(covidsm1 ~ age_g + sexvar + race + educag + incomg1 + hlthpl1 + smoker3 + bmi + menthlth + cvax, 
      data = brfss_c,
      method = "glm", 
      family = binomial(link = "logit"), 
      trControl = trainControl(method = "cv", 
                               number = 10, 
                               classProbs = TRUE, 
                               savePredictions = TRUE),
      weights = brfss_c$classwt)
# Deriving Sensitivity, Specificity, Accuracy, Kappa
confusionMatrix(lc_log$pred$pred, lc_log$pred$obs)
# Deriving Variable Importance
lc_log_vi <- varImp(lc_log$finalModel)
lc_log_vi[order(-lc_log_vi$Overall), , drop = FALSE]
# Seeing Model Coefficients and Odds Ratios
exp(coef(lc_log$finalModel))
summary(lc_log$finalModel)
```


```{r Naive Bayes}
# Naive Bayes
set.seed(1234)
lc_naive <- train(covidsm1 ~ age_g + sexvar + race + educag + incomg1 + hlthpl1 + smoker3 + bmi + menthlth + cvax, 
      data = brfss_c,
      method = "naive_bayes",
      tuneGrid = data.frame(laplace = 0, 
                            usekernel = FALSE, 
                            adjust = FALSE),
      trControl = trainControl(method = "cv", 
                               number = 10, 
                               classProbs = TRUE, 
                               savePredictions = TRUE),
      weights = brfss_c$classwt)
# Deriving Sensitivity, Specificity, Accuracy, Kappa
confusionMatrix(lc_naive$pred$pred, lc_naive$pred$obs)
# Naive Bayes is incompatible with varImp()
```


```{r Classification Tree}
# Setting up cp tuning grid
set.seed(1234)
lc_tree_cp <- rpart(covidsm1 ~ age_g + sexvar + race + educag + incomg1 + hlthpl1 + smoker3 + bmi + menthlth + cvax,
                    data = brfss_c,
                    method = "class",
                    parms = list(split = "gini"),
                    control = rpart.control(minsplit = 30, minbucket = 10), cp = 0,
                    weights = brfss_c$classwt)
tg_lc <- data.frame(cp = lc_tree_cp$cptable[,1]) 
# Classification Tree 
set.seed(1234)
lc_tree <- train(covidsm1 ~ age_g + sexvar + race + educag + incomg1 + hlthpl1 + smoker3 + bmi + menthlth + cvax,
                 data = brfss_c,
                 method = "rpart",
                 parms = list(split = "gini"),
                 control = rpart.control(minsplit = 30, minbucket = 10), 
                 tuneGrid = tg_lc,
                 trControl = trainControl(method = "cv",
                                          number = 10,
                                          selectionFunction = "oneSE",
                                          classProbs = TRUE,
                                          savePredictions = TRUE),
                 weights = brfss_c$classwt)
lc_tree
rpart.plot(lc_tree$finalModel)
# Deriving Sensitivity, Specificity, Accuracy, Kappa
confusionMatrix(lc_tree$pred$pred, lc_tree$pred$obs)
# Deriving Variable Importance
lc_tree_vi <- varImp(lc_tree$finalModel)
lc_tree_vi[order(-lc_tree_vi$Overall), , drop = FALSE]
```


```{r Random Forest}
# Random Forest
set.seed(1234)
lc_rf <-randomForest(covidsm1 ~ age_g + sexvar + race + educag + incomg1 + hlthpl1 + smoker3 + bmi + menthlth + cvax,
                     data=brfss_c,
                     classwt=class_weight,
                     importance=TRUE)
lc_rf
# Since the OOB Error Rate does appear to flatten, 500 trees is sufficient
plot(lc_rf$err.rate[, 1], col = "maroon", xlab = "Number of Trees", ylab = "OOB Error Rate")
# Calculating Kappa
oob_predictions <- lc_rf$votes
oob_predicted_classes <- factor(
  ifelse(oob_predictions[, 1] > 0.5, "Pos", "Neg")
)
oob_predicted_classes <- relevel(oob_predicted_classes, ref="Pos")
actual_classes <- brfss_c$covidsm1
conf_matrix <- table(Predicted = oob_predicted_classes, Actual = actual_classes)
kappa(conf_matrix)
#Deriving Variable Importance
lc_rf_vi <- varImp(lc_rf)
lc_rf_vi[order(-lc_rf_vi$Neg), , drop = FALSE]
varImpPlot(lc_rf, main="Variable Importance")
```


```{r results='hide'}
# Setting up tuning grid
tg_boostclass <- expand.grid(n.trees = c(5, 10, 15, 20, 25, 30, 35, 40, 45, 50),
                             interaction.depth = 1:6, 
                             shrinkage = c(0.2, 0.25, 0.3, 0.35, 0.4, 0.45), 
                             n.minobsinnode = 10)
# Boosted Tree
set.seed(1234)
lc_boost <- train(covidsm1 ~ age_g + sexvar + race + educag + incomg1 + hlthpl1 + smoker3 + bmi + menthlth + cvax,
               data = brfss_c,
               method = "gbm", 
               bag.fraction = 0.5, 
               tuneGrid = tg_boostclass,
               trControl = trainControl(method = "cv", 
                                        number = 10, 
                                        classProbs = TRUE,
                                        savePredictions = TRUE),
               weights = brfss_c$classwt)
```


```{r Boosted Tree}
# Deriving Sensitivity, Specificity, Accuracy, Kappa
confusionMatrix(lc_boost$pred$pred, lc_boost$pred$obs)
# Deriving Variable Importance
lc_boost_vi <- varImp(lc_boost$finalModel)
lc_boost_vi[order(-lc_boost_vi$Overall), , drop = FALSE]
summary(lc_boost$finalModel, 
        method = relative.influence, 
        normalize = TRUE,
        las = 2, 
        order=TRUE, 
        cBars=5)
plot(lc_boost$finalModel, i = "menthlth")
plot(lc_boost$finalModel, i = "educag4")
plot(lc_boost$finalModel, i = "sexvar2")
```
